Live Functional Programming with Typed Holes

### 2018-11-10 abstract.tex
format tweaking after adding copyright information

\begin{abstract}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% PAR 1: Live programming environments --- Problem statement
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

Live programming environments aim to provide programmers (and sometimes audiences) 
with continuous feedback about a program's dynamic behavior as it is being edited. 
%
The problem is that programming languages typically assign dynamic meaning only 
 to programs that are \emph{complete}, i.e. syntactically well-formed and free
of type errors. Consequently,    
live feedback presented to the programmer exhibits temporal or perceptive gaps. 
% In some cases, these gaps persist over a substantial length of time.
%
% Defining a semantics for incomplete functional programs would help bridge the
% gap between well-formed states and allow live programming environments to
% provide continuous, uninterrupted, and meaningful feedback.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% PAR 2: This paper's approach: New (dynamic) semantics for incomplete programs
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

This paper confronts this ``{gap problem}'' from type-theoretic first principles by developing 
\emph{a dynamic semantics for incomplete functional programs}, 
starting from the static semantics for incomplete functional programs developed in our recent work on \Hazelnut. 
%
We model incomplete functional programs as expressions with \emph{holes}, 
with empty holes standing for missing expressions or types, and  non-empty holes 
operating as membranes around static and dynamic type inconsistencies. 
%
Rather than aborting when evaluation encounters any of these holes as in
some existing systems, evaluation proceeds around holes,
tracking the 
closure around each hole instance as it flows through the remainder of the program. Editor services can use the information in these hole closures 
to help the programmer develop and confirm their mental model of the behavior of the complete portions of the program as they decide how to fill the remaining holes. 
Hole closures also enable a \emph{fill-and-resume} operation that 
avoids the need to restart evaluation after edits that amount to hole filling. 
Formally, the semantics borrows machinery from both gradual type theory (which supplies the basis for handling unfilled 
type holes) and contextual modal type theory (which supplies a
logical basis for hole closures), combining these and developing additional machinery necessary 
to continue evaluation past holes while maintaining type safety. We have mechanized the metatheory of the core calculus, called \HazelnutLive{}, using the Agda proof assistant.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% PAR 3: We have an implementation, and have proofs
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

% , including 
% (1) a type safety property that accounts for expression holes (i.e. free metavariables, 
% from the perspective of CMTT) and cast failure holes (which provide a novel), and (2) a commutativity property that establishes that the
% ``edit-and-resume'' feature is sound.
%
%
We have also implemented these ideas into the \Hazel programming environment. The implementation inserts holes automatically, following the \Hazelnut edit action calculus, to guarantee that every editor state has some (possibly incomplete)
type.
Taken together with this paper's type safety property, the 
% The 
result is  
a proof-of-concept live programming environment where rich dynamic feedback 
is truly available without gaps, i.e. for every reachable editor state.
% i.e. \emph{every edit state is guaranteed to have
% non-trivial static and dynamic meaning}.

>>>>

\begin{abstract}
Live programming environments aim to provide programmers (and sometimes audiences) 
with continuous feedback about a program's dynamic behavior as it is being edited. 
The problem is that programming languages typically assign dynamic meaning only 
to programs that are \emph{complete}, i.e. syntactically well-formed and free
of type errors. Consequently,    
live feedback presented to the programmer exhibits temporal or perceptive gaps.

This paper confronts this ``{gap problem}'' from type-theoretic first principles by developing 
\emph{a dynamic semantics for incomplete functional programs}, 
starting from the static semantics for incomplete functional programs developed in recent work on \Hazelnut. 
We model incomplete functional programs as expressions with \emph{holes}, 
with empty holes standing for missing expressions or types, and  non-empty holes 
operating as membranes around static and dynamic type inconsistencies. 
Rather than aborting when evaluation encounters any of these holes as in
some existing systems, evaluation proceeds around holes,
tracking the 
closure around each hole instance as it flows through the remainder of the program. Editor services can use the information in these hole closures 
to help the programmer develop and confirm their mental model of the behavior of the complete portions of the program as they decide how to fill the remaining holes. 
Hole closures also enable a \emph{fill-and-resume} operation that 
avoids the need to restart evaluation after edits that amount to hole filling. 
Formally, the semantics borrows machinery from both gradual type theory (which supplies the basis for handling unfilled 
type holes) and contextual modal type theory (which supplies a
logical basis for hole closures), combining these and developing additional machinery necessary 
to continue evaluation past holes while maintaining type safety. We have mechanized the metatheory of the core calculus, called \HazelnutLive{}, using the Agda proof assistant.

We have also implemented these ideas into the \Hazel programming environment. The implementation inserts holes automatically, following the \Hazelnut edit action calculus, to guarantee that every editor state has some (possibly incomplete)
type.
Taken together with this paper's type safety property, the 
result is  
a proof-of-concept live programming environment where rich dynamic feedback 
is truly available without gaps, i.e. for every reachable editor state.

### 2018-11-10 calculus.tex
formatting tweaks after inserting the ACM copyright stuff
tweaks to spacing
make section titles all caps again (blah)
most recent acmart and bibliography files
format tweaking after adding copyright information

% !TEX root = hazelnut-dynamics.tex
\newcommand{\calculusSec}{Hazelnut Live}
\section{\protect\calculusSec}
\label{sec:calculus}
% \def \TirNameStyle \Vtexttt{#1}{{#1}}

\input{fig-syntax}
\input{fig-bidirectional-typing}

@@

\parahead{Overview} The syntax of the core calculus given in
Fig.~\ref{fig:hazelnut-live-syntax} consists of types and expressions with
holes.  We distinguish between {external} expressions, $e$, and {internal}
expressions, $d$.  External expressions correspond to programs as entered
by the programmer (see Sec.~\ref{sec:intro} for discussion of implicit,
manual, semi-automated and fully automated hole entry methods).  Each
well-typed external expression (see Sec.~\ref{sec:external-statics} below)
elaborates to a well-typed internal expression (see
Sec.~\ref{sec:elaboration}) before it is evaluated (see
Sec.~\ref{sec:evaluation}).  We take this approach, notably also taken in
the ``redefinition'' of Standard ML by \citet{Harper00atype-theoretic},
because (1) the external language supports type inference and explicit type
ascriptions, $\hexp : \htau$, but it is formally simpler to eliminate
ascriptions and specify a type assignment system when defining the dynamic
semantics; and (2) we need additional syntactic machinery during evaluation
for tracking hole closures and dynamic type casts.  This machinery is
inserted by elaboration, rather than entered explicitly by the programmer.
In this regard, the internal language is analogous to the cast calculus in
the gradually typed lambda calculus
\cite{DBLP:conf/snapl/SiekVCB15,Siek06a}, though as we will see the
\HazelnutLive internal language goes beyond the cast calculus in several
respects. We have mechanized these formal developments using the Agda proof
assistant \cite{norell:thesis,norell2009dependently} (see
Sec.~\ref{sec:agda-mechanization}). Rule names in this section,
e.g. \rulename{SVar}, correspond to variable names used in the
mechanization. The \Hazel implementation substantially follows the formal
specification of \Hazelnut (for the editor component) and \HazelnutLive
(for the evaluator), and we can formally state a continuity invariant for a
putative combined calculus (see Sec.~\ref{sec:implementation}).

@@

@@

Rule \rulename{SVar} synthesizes the corresponding type from $\hGamma$.
For the sake of exposition, \HazelnutLive includes ``half-annotated'' lambdas, $\halam{x}{\htau}{\hexp}$, in addition to the unannotated lambdas, $\hlam{x}{\hexp}$, from \Hazelnut.
%
Half-annotated lambdas may appear in synthetic position according to Rule \rulename{SLam}, which is standard \cite{Chlipala:2005da}.
%
Unannotated lambdas may only appear where the expected type is known to be either an arrow type or the hole type, which is treated as if it were $\tarr{\tehole}{\tehole}$.\footnote{A system supporting ML-style type reconstruction \cite{damas1982principal} might also include a synthetic rule for unannotated lambdas, e.g. as outlined by \citet{DBLP:conf/icfp/DunfieldK13}, but we stick to this simpler ``Scala-style'' local type inference scheme in this paper \cite{Pierce:2000ve,Odersky:2001lb}.} 
%
To avoid the need for separate rules for these two cases, Rule \rulename{ALam} uses the matching relation $\arrmatch{\htau}{\tarr{\htau_1}{\htau_2}}$ defined in \Figref{fig:arrmatch}, which produces the matched arrow type $\tarr{\tehole}{\tehole}$ given the hole type, and operates as the identity on arrow types \cite{DBLP:conf/snapl/SiekVCB15,DBLP:conf/popl/GarciaC15}.
%
% \Secref{sec:related} dicusses how \HazelnutLive might be enriched with
% with ML-style type reconstruction~\cite{damas1982principal}, perhaps via
% the approach outlined by~\citet{DBLP:conf/icfp/DunfieldK13}.
%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% To Cyrus from Matt:
%
% Why say the following here? --- It's discussing a different design that we didn't pursue here.  We should move such discussion to related work.
%
%Note that
%%

The rule governing function application, Rule \rulename{SAp}, similarly treats an expression of hole type in function position as if it were of type $\tarr{\tehole}{\tehole}$ using the same matching relation.

@@

To recap, the result of elaboration is a well-typed internal expression
with appropriately initialized hole closures and casts.  This section
specifies the dynamic semantics of \HazelnutLive as a ``small-step''
transition system over internal expressions equipped with a meaningful
notion of type safety even for incomplete programs, i.e. expressions typed
under a non-empty hole context, $\Delta$.  We establish that evaluation
does not stop immediately when it encounters a hole, nor when a cast fails,
by precisely characterizing when evaluation \emph{does} stop. In the case
of complete programs, we also recover the familiar statements of progress
and preservation for the simply typed lambda calculus.

It is perhaps worth stating at the outset that a dynamic semantics equipped
with these properties does not simply ``fall out'' from the observations
made above that (1) empty hole closures correspond to metavariable closures
from CMTT \cite{Nanevski2008} and (2) casts also arise in gradual type
theory \cite{DBLP:conf/snapl/SiekVCB15}.
%
We say more in Sec.~\ref{sec:relatedWork}.
%

@@

\subsubsection{Application and Substitution}
%
Rule \rulename{ITLam} in Fig.~\ref{fig:instruction-transitions} defines the
standard beta reduction transition.
%
The bracketed premises of the form $\maybePremise{\isFinal{\dexp}}$ in
Fig.~\ref{fig:instruction-transitions}-\ref{fig:step} may be \emph{included}
to specify an eager, left-to-right evaluation strategy, or \emph{excluded} to
leave the evaluation strategy and order unspecified.
%
In our metatheory, we exclude these premises, both for
the sake of generality, and to support the specification of the fill-and-resume operation~(see \Secref{sec:resumption}).

@@

\input{fig-dynamics-contexts}

@@

The supplemental material includes our Agda
mechanization  \cite{norell2009dependently,norell:thesis,Aydemir:2005fk}
of the semantics and metatheory of \HazelnutLive,
%including proofs of all of the
including all of the theorems stated above and necessary lemmas.
%
%We choose the
%(as did the mechanization of \Hazelnut by \citet{popl-paper}, though only a few definitions are common).
%
%Agda is a good choice because it is designed to explicitly communicate a proof's structure, as is our goal, rather than relying on proof automation.
%
%Agda itself was also an inspiration for this work because it supports holes, albeit in a more limited form than described here (cf. Sec.~\ref{sec:intro}).
%
%
Our approach is standard: we model judgements as
inductive datatypes, and rules as dependently typed constructors of these judgements.
%
We adopt Barendregt's convention for bound variables \cite{urban,barendregt84:_lambda_calculus} and hole names, and avoid certain other complications related to substitution by enforcing the requirement that all bound variables in a term are unique when convenient (this requirement can always be discharged by alpha-variation). We encode typing
contexts and hole contexts using metafunctions.
To support this encoding choice, we postulate function extensionality (which is independent of Agda's axioms) \cite{awodey2012inductive}. We encode finite substitutions as an inductive datatype with a base case representing the identity substitution and an inductive case that records a single substitution. Every finite substitution can be represented this way. This makes it easier for Agda to see why certain inductions that we perform are well-founded.  
The documentation provided with the mechanization has more details.

@@

\newcommand{\commutativitySec}{A Contextual Modal Interpretation of Fill-and-Resume}
\section{\protect\commutativitySec}
\label{sec:resumption}

>>>>

% !TEX root = hazelnut-dynamics.tex
\newcommand{\calculusSec}{Hazelnut Live}
\section{\calculusSec}
\label{sec:calculus}
% \def \TirNameStyle \Vtexttt{#1}{{#1}}

\input{fig-syntax}

@@

\noindent
\parahead{Overview} The syntax of the core calculus given in
Fig.~\ref{fig:hazelnut-live-syntax} consists of types and expressions with
holes.  We distinguish between {external} expressions, $e$, and {internal}
expressions, $d$.  External expressions correspond to programs as entered
by the programmer (see Sec.~\ref{sec:intro} for discussion of implicit,
manual, semi-automated and fully automated hole entry methods).  Each
well-typed external expression (see Sec.~\ref{sec:external-statics} below)
elaborates to a well-typed internal expression (see
Sec.~\ref{sec:elaboration}) before it is evaluated (see
Sec.~\ref{sec:evaluation}).  We take this approach, notably also taken in
the ``redefinition'' of Standard ML by \citet{Harper00atype-theoretic},
because (1) the external language supports type inference and explicit type
ascriptions, $\hexp : \htau$, but it is formally simpler to eliminate
ascriptions and specify a type assignment system when defining the dynamic
semantics; and (2) we need additional syntactic machinery during evaluation
for tracking hole closures and dynamic type casts.  This machinery is
inserted by elaboration, rather than entered explicitly by the programmer.
In this regard, the internal language is analogous to the cast calculus in
the gradually typed lambda calculus
\cite{DBLP:conf/snapl/SiekVCB15,Siek06a}, though as we will see the
\HazelnutLive internal language goes beyond the cast calculus in several
respects. We have mechanized these formal developments using the Agda proof
assistant \cite{norell:thesis,norell2009dependently} (see
Sec.~\ref{sec:agda-mechanization}). Rule names in this section,
e.g. \rulename{SVar}, correspond to variables from the
mechanization. The \Hazel implementation substantially follows the formal
specification of \Hazelnut (for the editor) and \HazelnutLive
(for the evaluator). We can formally state a continuity invariant for a
putative combined calculus (see Sec.~\ref{sec:implementation}).

@@

\input{fig-bidirectional-typing}

@@

Rule \rulename{SVar} synthesizes the corresponding type from $\hGamma$.
For the sake of exposition, \HazelnutLive includes ``half-annotated'' lambdas, $\halam{x}{\htau}{\hexp}$, in addition to the unannotated lambdas, $\hlam{x}{\hexp}$, from \Hazelnut.
%
Half-annotated lambdas may appear in synthetic position according to Rule \rulename{SLam}, which is standard \cite{Chlipala:2005da}.
%
Unannotated lambdas may only appear where the expected type is known to be either an arrow type or the hole type, which is treated as if it were $\tarr{\tehole}{\tehole}$.\footnote{A system supporting ML-style type reconstruction \cite{damas1982principal} might also include a synthetic rule for unannotated lambdas, e.g. as outlined by \citet{DBLP:conf/icfp/DunfieldK13}, but we stick to this simpler ``Scala-style'' local type inference scheme in this paper \cite{Pierce:2000ve,Odersky:2001lb}.} 
%
To avoid the need for separate rules for these two cases, Rule \rulename{ALam} uses the matching relation $\arrmatch{\htau}{\tarr{\htau_1}{\htau_2}}$ defined in \Figref{fig:arrmatch}, which produces the matched arrow type $\tarr{\tehole}{\tehole}$ given the hole type, and operates as the identity on arrow types \cite{DBLP:conf/snapl/SiekVCB15,DBLP:conf/popl/GarciaC15}. The rule governing function application, Rule \rulename{SAp}, similarly treats an expression of hole type in function position as if it were of type $\tarr{\tehole}{\tehole}$ using the same matching relation.
%
% \Secref{sec:related} dicusses how \HazelnutLive might be enriched with
% with ML-style type reconstruction~\cite{damas1982principal}, perhaps via
% the approach outlined by~\citet{DBLP:conf/icfp/DunfieldK13}.
%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% To Cyrus from Matt:
%
% Why say the following here? --- It's discussing a different design that we didn't pursue here.  We should move such discussion to related work.
%
%Note that
%%

@@

To recap, the result of elaboration is a well-typed internal expression
with appropriately initialized hole closures and casts.  This section
specifies the dynamic semantics of \HazelnutLive as a ``small-step''
transition system over internal expressions equipped with a meaningful
notion of type safety even for incomplete programs, i.e. expressions typed
under a non-empty hole context, $\Delta$.  We establish that evaluation
does not stop immediately when it encounters a hole, nor when a cast fails,
by precisely characterizing when evaluation \emph{does} stop. In the case
of complete programs, we also recover the familiar statements of preservation
and progress for the simply typed lambda calculus.

% It is perhaps worth stating at the outset that a dynamic semantics equipped
% with these properties does not simply ``fall out'' from the observations
% made above that (1) empty hole closures correspond to metavariable closures
% from CMTT \cite{Nanevski2008} and (2) casts also arise in gradual type
% theory \cite{DBLP:conf/snapl/SiekVCB15}.
%
% We say more in Sec.~\ref{sec:relatedWork}.
%

@@

\subsubsection{Application and Substitution}
\input{fig-dynamics-contexts}
%
Rule \rulename{ITLam} in Fig.~\ref{fig:instruction-transitions} defines the
standard beta reduction transition.
%
The bracketed premises of the form $\maybePremise{\isFinal{\dexp}}$ in
Fig.~\ref{fig:instruction-transitions}-\ref{fig:step} may be \emph{included}
to specify an eager, left-to-right evaluation strategy, or \emph{excluded} to
leave the evaluation strategy and order unspecified.
%
In our metatheory, we exclude these premises, both for
the sake of generality, and to support the specification of the fill-and-resume operation~(see \Secref{sec:resumption}).

@@

@@

The archived artifact includes our Agda
mechanization  \cite{norell2009dependently,norell:thesis,Aydemir:2005fk}
of the semantics and metatheory of \HazelnutLive,
%including proofs of all of the
including all of the theorems stated above and necessary lemmas.
%
%We choose the
%(as did the mechanization of \Hazelnut by \citet{popl-paper}, though only a few definitions are common).
%
%Agda is a good choice because it is designed to explicitly communicate a proof's structure, as is our goal, rather than relying on proof automation.
%
%Agda itself was also an inspiration for this work because it supports holes, albeit in a more limited form than described here (cf. Sec.~\ref{sec:intro}).
%
%
Our approach is standard: we model judgements as
inductive datatypes, and rules as dependently typed constructors of these judgements.
%
We adopt Barendregt's convention for bound variables \cite{urban,barendregt84:_lambda_calculus} and hole names, and avoid certain other complications related to substitution by enforcing the requirement that all bound variables in a term are unique when convenient (this requirement can always be discharged by alpha-variation). We encode typing
contexts and hole contexts using metafunctions.
To support this encoding choice, we postulate function extensionality (which is independent of Agda's axioms) \cite{awodey2012inductive}. We encode finite substitutions as an inductive datatype with a base case representing the identity substitution and an inductive case that records a single substitution. Every finite substitution can be represented this way. This makes it easier for Agda to see why certain inductions that we perform are well-founded.  
The documentation provided with the mechanization has more details.

@@

\newcommand{\commutativitySec}{A Contextual Modal Interpretation of Fill-and-Resume}
\section{\commutativitySec}
\label{sec:resumption}

### 2018-11-10 discussion.tex
make section titles all caps again (blah)
format tweaking after adding copyright information

% !TEX root = hazelnut-dynamics.tex
\newcommand{\discussionSection}{Conclusion}
\section{\protect\discussionSection} % don't like the all-caps thing that the template does, so protecting it from that
\label{sec:discussion}

@@

\noindent
%
Weinberg's sentiment applies to countless situations that programmers face as they work to develop and critically investigate the mental model of the program they are writing---at the very moment where rich feedback might be most helpful, e.g. when there is an error in the program or when the programmer is unsure how to fill a hole, the computer often has comparatively little feedback to offer (perhaps just a parser error message, or an explanation of one type error). 
%
% Efforts to develop tools that help programmers explore the behavior of parts of their program interactively, i.e. live programming tools, have been increasingly common in the literature on programming languages, software engineering, and human-computer interaction. 
%
% Yet much of this work has assumed that the programmer has already written a formally complete program, or relies on \emph{ad hoc} heuristics to handle some incomplete states. 
%
It is our hope that the well-behaved  type-theoretic foundations developed here will enable not only \Hazel but live programming tools of a wide variety of other designs to further narrow the temporal and perceptive gap and provide meaningful feedback to programmers at these important   moments.

\newcommand{\acksSection}{Acknowledgments}
\section*{\protect\acksSection} 
\label{sec:acks}

We thank Brigitte Pientka, Jonathan Aldrich, Joshua Sunshine, Michael Hilton, Claire LeGoues, Conor McBride, 
the participants of the PL reading group at UChicago,
the participants of the LIVE 2017 and Off the Beaten Track (OBT)
2017 workshops, and the anonymous referees for their insights and feedback on various iterations of this work.
%
This material was supported by a gift from
Facebook, from the National Science Foundation under grant
numbers CCF-1619282, SHF-1814900 and SHF-1817145, and from AFRL and DARPA under agreement \#FA8750-16-2-0042. 
The U.S. Government is authorized to reproduce and distribute reprints for 
Governmental purposes notwithstanding any copyright notation thereon.
%
Any opinions, findings, and conclusions or recommendations expressed
in this material are those of the authors and do not necessarily
reflect the views of Facebook, NSF, DARPA, AFRL or the U.S. Government. 
%
\clearpage
% \todo{discussion}

>>>>

% !TEX root = hazelnut-dynamics.tex
\newcommand{\discussionSection}{Conclusion}
\section{\discussionSection} % don't like the all-caps thing that the template does, so protecting it from that
\label{sec:discussion}

@@

\noindent
%
Weinberg's sentiment applies to countless situations that programmers face as they work to develop and critically investigate the mental model of the program they are writing---at the very moment where rich feedback might be most helpful, e.g. when there is an error in the program or when the programmer is unsure how to fill a hole, the computer often has comparatively little feedback to offer (perhaps just a parser error message, or an austere explanation of a type error). 
%
% Efforts to develop tools that help programmers explore the behavior of parts of their program interactively, i.e. live programming tools, have been increasingly common in the literature on programming languages, software engineering, and human-computer interaction. 
%
% Yet much of this work has assumed that the programmer has already written a formally complete program, or relies on \emph{ad hoc} heuristics to handle some incomplete states. 
%
It is our hope that the well-behaved  type-theoretic foundations developed here will enable not only \Hazel but live programming tools of a wide variety of other designs to further narrow the temporal and perceptive gap and provide meaningful feedback to programmers at these important   moments.

\newcommand{\acksSection}{Acknowledgments}
\section*{\acksSection} 
\label{sec:acks}

We thank Brigitte Pientka, Jonathan Aldrich, Joshua Sunshine, Michael Hilton, Claire LeGoues, Conor McBride, 
the participants of the PL reading group at UChicago,
the participants of the LIVE 2017 and Off the Beaten Track (OBT)
2017 workshops, and the anonymous referees for their insights and feedback on various iterations of this work.
%
This material was supported by a gift from
Facebook, from the National Science Foundation under grant
numbers CCF-1619282, SHF-1814900 and SHF-1817145, and from AFRL and DARPA under agreement \#FA8750-16-2-0042. 
 The U.S. Government is authorized to reproduce and distribute reprints for 
Governmental purposes notwithstanding any copyright notation thereon.
%
Any opinions, findings, and conclusions or recommendations expressed
in this material are those of the authors and do not necessarily
reflect the views of Facebook, NSF, DARPA, AFRL or the U.S. Government. 
%
% \clearpage
% \todo{discussion}

### 2018-11-10 examples.tex
tweaks to spacing
format tweaking after adding copyright information

% !TEX root = hazelnut-dynamics.tex
%% \newcommand{\examplesSec}{Live Functional Programming in Hazel, By Example}
%% \newcommand{\examplesSec}{Live Programming in \HazelnutLive, By Example}
\newcommand{\examplesSec}{Live Programming in Hazel}
\section{\protect\examplesSec} % don't like the all-caps thing that the template does, so protecting it from that
\label{sec:examples}

@@

Let us start with an example-driven overview of this paper's approach in \Hazel, a live programming environment being developed by \citet{HazelnutSNAPL}. The \Hazel user interface is based roughly on IPython/Jupyter \cite{PER-GRA:2007}, with a result appearing below each cell that contains an expression, and the \Hazel language is tracking toward feature parity with \Elm~(\url{elm-lang.org}) \cite{czaplicki2012elm,Elm}, a popular pure functional programming language similar to ``core ML'', with which we assume familiarity. \Hazel is intended initially for use by students and instructors in introductory functional programming courses (where \Elm~ has been successful \cite{DBLP:journals/corr/abs-1805-05125}). 
For the sake of 
exposition, we have post-processed the screenshots in this section after generating them in \Hazel to make use of  ``syntactic and semantic sugar'' from \Elm~that was not available in \Hazel (which, as of this writing, implements little more than the language features described in Sec.~\ref{sec:calculus} and the appendix). These conveniences are orthogonal to the contributions of this paper; all of the user interface features demonstrated in this section have been implemented and all of the computations can be expressed using standard ``encoding tricks''.

>>>>

% !TEX root = hazelnut-dynamics.tex
\newcommand{\examplesSec}{Live Programming in Hazel}
\section{\examplesSec} 
\label{sec:examples}

@@

Let us start with an example-driven overview of this paper's approach in \Hazel, a live programming environment being developed by \citet{HazelnutSNAPL}. The \Hazel user interface is based roughly on IPython/Jupyter \cite{PER-GRA:2007}, with a result appearing below each cell that contains an expression, and the \Hazel language is tracking toward feature parity with \Elm~(\url{elm-lang.org}) \cite{czaplicki2012elm,Elm}, a popular pure functional programming language similar to ``core ML'', with which we assume familiarity. \Hazel is intended initially for use by students and instructors in introductory functional programming courses (where \Elm~ has been successful \cite{DBLP:journals/corr/abs-1805-05125,zhang2018graphics}).

For the sake of 
exposition, we have post-processed the screenshots in this section after generating them in \Hazel to make use of  ``syntactic and semantic sugar'' from \Elm~that was not available in \Hazel (which, as of this writing, implements little more than the language features described in Sec.~\ref{sec:calculus} and the appendix). These conveniences are orthogonal to the contributions of this paper; all of the user interface features demonstrated in this section have been implemented and all of the computations can be expressed using standard ``encoding tricks''.

### 2018-11-10 fig-grades-example.tex
missed a file a few commits ago
format tweaking after adding copyright information

\begin{figure}[t]
\begin{subfigure}[t]{\textwidth}
\centering
\includegraphics[width=0.8\textwidth,interpolate=false]{images/grades-cell-mockup.png}
\vspace{-3px}
\caption{Evaluating an incomplete functional program past the first hole}
\label{fig:grades-cell-mockup}
\end{subfigure}

@@

\begin{subfigure}[t]{\textwidth}
\centering
\includegraphics[width=0.29\textwidth,interpolate=false,valign=c]{images/grades-sidebar-1.png}
~${}^\blacktriangleright$
\includegraphics[width=0.29\textwidth,interpolate=false,valign=c]{images/grades-sidebar-2.png}
~${}^\blacktriangleright$
\includegraphics[width=0.29\textwidth,interpolate=false,valign=c]{images/grades-sidebar-3.png}
\caption{The live context inspector communicates relevant static \emph{and} dynamic information about variables in scope.}
\label{fig:grades-sidebar}
\end{subfigure}
% %% TODO once the code above is removed, scale up the screenshots
% \includegraphics[scale=0.20]{images/hazel-placeholder-0.png}

>>>>

\begin{figure}[t]
\begin{subfigure}[t]{\textwidth}
\centering
\includegraphics[width=\textwidth,interpolate=false]{images/grades-cell-mockup.png}
\vspace{-10px}
\caption{Evaluating an incomplete functional program past the first hole}
\label{fig:grades-cell-mockup}
\end{subfigure}

@@

\begin{subfigure}[t]{\textwidth}
\centering
\includegraphics[width=0.31\textwidth,interpolate=false,valign=c]{images/grades-sidebar-1.png}
~${}^\blacktriangleright$
\includegraphics[width=0.31\textwidth,interpolate=false,valign=c]{images/grades-sidebar-2.png}
~${}^\blacktriangleright$
\includegraphics[width=0.31\textwidth,interpolate=false,valign=c]{images/grades-sidebar-3.png}
\caption{The live context inspector communicates relevant static \emph{and} dynamic information about variables in scope.}
\label{fig:grades-sidebar}
\end{subfigure}
% %% TODO once the code above is removed, scale up the screenshots
% \includegraphics[scale=0.20]{images/hazel-placeholder-0.png}

### 2018-11-10 hazelnut-dynamics.tex
small clarifications through sec 3, addressing R2's confusion about matching relation, and more editor states being meaningful; also changed the description of the mechanization to be consistent with our final artifact
formatting tweaks after inserting the ACM copyright stuff
tweak abstract to remove citations per the ACM formatting requirements
tweaks to spacing
format tweaking after adding copyright information

\PassOptionsToPackage{svgnames,dvipsnames,svgnames}{xcolor}
\newif\ifarxiv
\arxivfalse
\ifarxiv
\documentclass[acmsmall,nonacm]{acmart}
\else
%%% Note: arxiv does not want line numbers (they are detected somehow, and are not allowed)
\documentclass[acmsmall]{acmart}
% \settopmatter{printfolios=false,printccs=false,printacmref=false}
% \settopmatter{printccs=false,printacmref=false}
\fi

\ifarxiv
\newcommand{\appendixName}{appendix}
\else
\newcommand{\appendixName}{supplemental appendix}
\fi

@@

\makeatletter\if@ACM@journal\makeatother
%% Journal information (used by PACMPL format)
%% Supplied to authors by publisher for camera-ready submission
\acmJournal{PACMPL}
\acmVolume{1}
\acmNumber{1}
\acmArticle{1}
\acmYear{2019}
\acmMonth{1}
\acmDOI{10.1145/nnnnnnn.nnnnnnn}
\startPage{1}
\else\makeatother
%% Conference information (used by SIGPLAN proceedings format)
%% Supplied to authors by publisher for camera-ready submission
% \acmConference[]{ACM SIGPLAN Conference on Programming Languages}{January 01--03, 2017}{New York, NY, USA}

\acmYear{2019}
\acmISBN{978-x-xxxx-xxxx-x/YY/MM}
\acmDOI{10.1145/nnnnnnn.nnnnnnn}
\startPage{1}
\fi

%% Copyright information
%% Supplied to authors (based on authors' rights management selection;
%% see authors.acm.org) by publisher for camera-ready submission
\setcopyright{none}             %% For review submission
%\setcopyright{acmcopyright}
%\setcopyright{acmlicensed}
%\setcopyright{rightsretained}
%\copyrightyear{2017}           %% If different from \acmYear

@@

%% Title information
\title{Live Functional Programming with Typed Holes}         %% [Short Title] is optional;
                                        %% when present, will be used in
                                        %% header instead of Full Title.
% \titlenote{with title note}             %% \titlenote is optional;
                                        %% can be repeated if necessary;
                                        %% contents suppressed with 'anonymous'
% \subtitle{Subtitle}                     %% \subtitle is optional
% \subtitlenote{with subtitle note}       %% \subtitlenote is optional;
                                        %% can be repeated if necessary;
                                        %% contents suppressed with 'anonymous'

@@

%% Keywords
%% comma separated list
\keywords{live programming, gradual typing, contextual modal type theory, dynamic software updating, typed holes, structured editing}  %% \keywords is optional

>>>>

\PassOptionsToPackage{svgnames,dvipsnames,svgnames}{xcolor}
\newif\ifarxiv
\arxivtrue
\ifarxiv
\documentclass[acmsmall,screen,nonacm]{acmart}
\else
%%% Note: arxiv does not want line numbers (they are detected somehow, and are not allowed)
\documentclass[acmsmall,screen]{acmart}
% \settopmatter{printfolios=false,printccs=false,printacmref=false}
% \settopmatter{printccs=false,printacmref=false}
\fi

\ifarxiv
\newcommand{\appendixName}{appendix}
\else
\newcommand{\appendixName}{extended appendix}
\fi

@@

%% Copyright information
%% Supplied to authors (based on authors' rights management selection;
%% see authors.acm.org) by publisher for camera-ready submission
% \setcopyright{none}             %% For review submission
%\setcopyright{acmcopyright}
%\setcopyright{acmlicensed}
%\setcopyright{rightsretained}
%\copyrightyear{2017}           %% If different from \acmYear
\setcopyright{rightsretained} 
\acmJournal{PACMPL}
\acmYear{2019} \acmVolume{3} \acmNumber{POPL} \ifarxiv \acmArticle{1} \else \acmArticle{14} \fi \acmMonth{1} \acmPrice{}\acmDOI{10.1145/3290327}
\copyrightyear{2019}

@@

%% Title information
\title{Live Functional Programming with Typed Holes}         %% [Short Title] is optional;
\subtitle{Extended Version}
\subtitlenote{The original version of this article appears as ...}
                                        %% when present, will be used in
                                        %% header instead of Full Title.
% \titlenote{with title note}             %% \titlenote is optional;
                                        %% can be repeated if necessary;
                                        %% contents suppressed with 'anonymous'
% \subtitle{Subtitle}                     %% \subtitle is optional
% \subtitlenote{with subtitle note}       %% \subtitlenote is optional;
                                        %% can be repeated if necessary;
                                        %% contents suppressed with 'anonymous'

@@

%% Keywords
%% comma separated list
\keywords{live programming, gradual typing, contextual modal type theory, typed holes, structured editing}  %% \keywords is optional
